### 2020-04-04 Thesis/thesis.tex
Corrections.
Formatting/spell changes

\begin{example}
  In the category of sets, the \emph{nonempty powerset functor} $\powerset_+$ sends a set $A$ to the set of nonempty subsets of $A$.  
  This functor has the structure of a monad on $\Set$, since we have a natural transformation (union) from $\powerset_+\powerset_+A$ to $powerset_+A$ and a natural transformation (singleton) from $A$ to$ \powerset_+A$ that obey the diagrams given above.
\end{example}
\begin{example}
  Let $\M$ be a monoidal category and let $x$ be a monoid in $\M$.  
  The \emph{writer monad} $W_x$ on $\M$ is defined by $W_x y = y \tensor x$, with natural transformations
  \begin{mathpar}
    m_y \from y \tensor x \tensor x \to y \tensor x
    \and
    u_y \from y \to y \tensor x
  \end{mathpar}
  given by the monoid structure on $x$.

@@

  As an example, the definition of a monoidal closed category $\M$ is that it admits an adjunction between functors $\blank\tensor w$ and $w\implies\blank$ for any object $w$ of $M$.  
  if $\M$ is a monoidal closed category and $w$ is an object of $\M$, then the functor $S_w$ on $\M$ defined by
  \[
    S_w x = w \implies (x \tensor w)
    \]
  is the composition of these adjoint functors and therefore inherits the structure of a monad, known as the \emph{state monad}.
\end{example}
\begin{example}
  Another example that arises from an adjunction is the \emph{list monad} on $\Set$ that arises from the adjunction between the $\Set$ and the category of monoids.
  The underlying set of the free monoid on a set $A$ is the set $A^*$ of finite lists of elements of $A$, and the functor $A\mapsto A^*$ inherits a monoid structure where the multiplication $m_A\from (A^*)^* \to A^*$ concatenates a list of lists into a single list and the unit $u_a \from A \to A^*$ forms a list with a single element.  
\end{example}
\begin{example}
  A monad on $\oppcat\C$ is called a \emph{comonad} on $\C$.  
  The carrier of a comonad is still a functor $M\from \C\to\C$, but now the multiplication and unit are natural transformations $M\Rightarrow MM$ and $M\Rightarrow 1$, rather than the other way round.

  An adjunction $\adjunction\C LR\D$ gives rise to a comonad structure on $LR$ in much the same way as it gives rise to a monad structure on $RL$.  
  So, for example, we have the \emph{store comonad} $S_r'$ for any object $r$ of a monoidal closed category $\M$, given by
  \[
    S_r'x = (r \implies x) \tensor x\,.
    \]
\end{example}

@@

\subsection{Multiplicative natural transformations}

@@

The proof of Proposition \ref{pKleisli} essentially comes down to the following factorization result.  
If $f\from a \to b$ is a morphism in $\Kl_M$, then $f$ may be factorized as
\[
  f = a \xrightarrow{Jf}
  Mb \xrightarrow{\phi_b}
  b\,,
  \]
where we use `$f$' to refer both to the morphism $a\to b$ in $\Kl_M$ and to the underlying morphism $a \to Mb$ in $\C$.
Indeed, if we compute this composite inside $\C$, we get
\[
  a \xrightarrow{f}
  Mb \xrightarrow{u_{Mb}}
  MMb \xrightarrow{M\id}
  MMb \xrightarrow{m_b}
  Mb\,,
  \]
which is equal to $f$ by the coherence conditions on $m$ and $u$.
This means that the Kleisli category may be thought of as being freely generated from the original category $\C$ and a multiplicative natural transformation $\phi$.

\begin{example}
  The morphisms in the Kleisli category for the nonempty powerset monad $\powerset_+$ on $\Set$ are functions $A \to \powerset_+B$, which can be thought of as nondeterministic functional programs.  
  Given a set $A$, the morphism $\phi_A\from \powerset_+A \to A$ in $\Kl_{\powerset_+}$ can be interpreted as a `nondeterministic choice' function that accepts a nonempty set of elements of $A$ and nondeterministically chooses one of them.
  The factorization then means that the category is freely generated over $\C$ by these nondeterministic choice morphisms.
\end{example}
\begin{example}
  Let $\C$ be a Cartesian closed category and let $z$ be some fixed object of $\C$.  
  Then the Kleisli category for the reader monad $R_z$ on $\C$ is generated over $\C$ by a natural transformation $\phi_y\from (z \to y) \to y$.  
  By the enriched Yoneda lemma, such a natural transformation is always given by precomposition with some fixed morphism $\ask_z\from 1 \to z$.  
  This means that $\Kl_{R_z}$ is suitable for modelling any situation in which we are generally working in $\C$, but need the ability to request a value of type $z$ (for example, a config file, a piece of user input, a random number or something else that isn't being passed into the function in question).
  \label{ExReaderMonadKleisli}
\end{example}

@@

Having given an overview of the general theory of monads and Kleisli categories, we now examine the relationship between Kleisli categories and Full Abstraction.  From now till the end of the chapter, we fix an base Cartesian closed category $\G$ that admits a denotational semantics of Idealized Algol satisfying Computational Adequacy.  
In addition, we require that $\G$ can be interpreted as being enriched in algebraic partial orders, in such a way that any compact morphism between the denotation of IA types is the denotation of some IA term.
The prototypical example, of course, will be the category of games and visible strategies, but we will not exploit any properties of this model beyond the ones we have already mentioned, mentioning it only in examples where appropriate.

@@

From Proposition \ref{pKleisli}, we know that there is a distinguished natural transformation $\phi_A \from (X \to A) \to A$ in $\G_X$; in particular, we have a morphism
\[
  \phi = \Lambda(\lunit_X;\id_X);\phi_X
  \]
(where $\Lambda(\lunit_X;\id_X)\from 1 \to (X \to X)$ is the interpretation of the $\lambda$-term $\lambda x.x$), which will be the denotation of the term $\choose_X$.
Together with the existing denotational semantics of Idealized Algol within $\G$, this gives us an inductively defined denotational semantics of \IAX within $\G_X$.

@@

  \begin{figure}
    \begin{mathpar}
      \inferrule*{ }{\Gamma,s\ts c \converges_\epsilon c,s}
      \and
      \inferrule*{\Gamma,s \ts M \converges_u \lambda x.M',s' \\ \Gamma,s' \ts M'[N/x] \converges_v c,s''}{\Gamma,s \ts MN \converges_{u\cat v} c,s''}
      \and
      \inferrule*{\Gamma,s \ts M(\Y M) \converges_u c,s'}{\Gamma,s \ts \Y M \converges_u c,s'}
      \and
      \inferrule*{\Gamma,s\ts M \converges_u n,s'}{\Gamma,s\ts \suc M \converges_u n+1,s'}
      \\\and
      \inferrule*{\Gamma,s\ts M \converges_u n+1,s'}{\Gamma,s\ts \pred M \converges_u n,s'}
      \and
      \inferrule*{\Gamma,s\ts M \converges_u 0,s'}{\Gamma,s\ts \pred M \converges_u 0,s'}
      \and
      \inferrule*{\Gamma,s\ts M \converges_u \skipp,s' \\ \Gamma,s'\ts N \converges_v c,s''}{\Gamma,s \ts M;N \converges_{u\cat v} c,s''}
      \and
      \inferrule*{\Gamma,s\ts M \converges_u \true,s' \\ \Gamma,s' \ts N \converges_v c,s''}{\Gamma,s \ts \If M \Then N \Else P \converges_{u\cat v} c,s''}
      \and
      \inferrule*{\Gamma,s\ts M \converges_u \false,s' \\ \Gamma,s' \ts P \converges_v c,s''}{\Gamma,s \ts \If M \Then N \Else P \converges_{u\cat v} c,s''}
      \and
      \inferrule*{\Gamma,s\ts M \converges_u 0,s' \\ \Gamma,s' \ts N \converges_v c,s''}{\Gamma,s \ts \IfO M \Then N \Else P \converges_{u\cat v} c,s''}
      \and
      \inferrule*{\Gamma,s\ts M \converges_u n+1,s' \\ \Gamma,s' \ts P \converges_v c,s''}{\Gamma,s \ts \IfO M \Then N \Else P \converges_{u\cat v} c,s''}
      \and
      \inferrule*[right=$x\in\Gamma$]{\Gamma,s\ts E \converges_u n,s' \\ \Gamma,s' \ts V \converges_v x,s''}{\Gamma,s\ts V \gets E \converges_{u\cat v} \skipp,(s''\vert x \mapsto n)}
      \and
      \inferrule*[right={$s'(x)=n$}]{\Gamma,s\ts V \converges_u x,s'}{\Gamma,s\ts !V \converges_u n,s'}
      \and
      \inferrule*{\Gamma,x\from\Var,(s\vert x\mapsto 0)\ts M \converges_u c,(s'\vert x\mapsto n)}{\Gamma,s\ts \neww \lambda x.M \converges_u c,s'}
      \and
      \inferrule*{\Gamma,s\ts E \converges_u n,s' \\ \Gamma,s'\ts V \converges_v \mkvar W R,s'' \\ \Gamma,s'' \ts Wn \converges_w \skipp,s'''}
      {\Gamma,s \ts V\gets E \converges_{u\cat v\cat w} \skipp,s'''}
      \and
      \inferrule*{\Gamma,s\ts V \converges_u \mkvar W R,s' \\ \Gamma,s'\ts R \converges_v n,s''}{\Gamma,s\ts !V \converges_{u\cat v} n,s''}
      \\\and
      \inferrule*[right=$x\in X$]{ }{\Gamma,s\ts \ask_X \converges_x x,s}
    \end{mathpar}
    \caption[Operational semantics for \IAX]{Operational semantics for \IAX.  
    All the rules except the last one are deterministic and may be obtained from the corresponding rules of Idealized Algol by suitably annotating the $\converges$ relation with sequences from $X^*$.}
    \label{FigIaxOpSem}
  \end{figure}

  Closer examination of the rules in Figure \ref{FigIaxOpSem} reveals an alternative, indirect definition of the operational semantics of \IAX.
  Note that each rule from ordinary Idealized Algol takes the form
  \[
    \inferrule{\Gamma,s^{(0)}\ts M_1\converges c_1,s^{(1)} \\ \cdots \\ \Gamma,s^{(n-1)}\ts M_n\converges c_n,s^{(n)}}
    {\Gamma,s^{(0)} \ts M \converges c,s^{(n)}}\,,
    \]
  Here, we have interpreted each IA rule as an infinite scheme of rules ranging over the different terms $M_i,M$ that the rule can apply to.
  We first extend this rule to a rule for \IAX, by allowing the $M_i,M$ to range over terms of \IAX.
  We then replace the rule with the new rule
  \[
    \inferrule{\Gamma,s^{(0)}\ts M_1 \converges_{u_1} c_1,s^{(1)} \\ \cdots \\ \Gamma,s^{(n-1)}\ts M_n \converges_{u_n} c_n,s^{(n)}}
    {\Gamma,s^{(0)} \ts M \converges_{u_1 \cat \cdots \cat u_n} c,s^{(n)}}\,,
    \]
  to give us an operational rule for \IAX (if $n=0$, then we treat the empty string $\epsilon$ as the empty concatenation).
  Lastly, we add the rule for the new constant $\ask_X$:
  \[
    \inferrule*[right=$x\in X$]{ }{\Gamma,s \ts \ask_X \converges_x x,s}\,.
    \]
  This rule is the only nondeterministic one in our language, as well as being the only one in which the sequence annotating the $\converges$ symbol at the bottom is not formed from concatenating together the sequences on the top.
\end{definition}

\begin{example}
  If $X=\bC$, then, since $X$ has a single element $a$, a sequence $u=\underbrace{a\cdots a}_n$ of elements of $X$ may be identified with its length $n$.
  In this case, the language \IAX gives us a way to model time complexity, and the term $\ask_X$ may be considered as a subroutine $\sleep\from \com$ whose semantics is to wait for some fixed period of time before continuing.  
  In this case, 
  \[
    \Gamma,s\ts M \converges_n c,s'
    \]
  is interpreted to say that `starting with the store $s$, $M$ converges to $c$ in time $n$, leaving store $s'$'.
\end{example}
\begin{example}
  If $X=\bN$, we can interpret the language \IAX as giving us a way to model user input.  
  When we call $\ask_\bN$, we are asking the user to provide us as a string of text (which we interpret as a binary string and hence as a natural number).  
  Then we interpret the judgement
  \[
    \Gamma,s\ts M \converges_u c,s'
    \]
  as saying that `starting with the store $s$, and the sequence $u$ of user input, $M$ converges to $c$, leaving store $s'$'.
\end{example}
\begin{example}
  If $X\in\{\bB,\bN\}$, then the language \IAX gives us a way to model nondeterminism, where $\ask_X$ behaves as a \emph{nondeterministic oracle}; i.e., a device that nondeterministically returns an element of $X$.

@@

\begin{definition}
  We inductively define terms in context $\tr_u\from \nat \to X$ of \emph{ordinary deterministic} Idealized Algol for each $u\in X^*$ as follows.
  \begin{mathpar}
    \tr_\epsilon = \lambda n.\Omega
    \and
    \tr_{xu} = \lambda n.\neww (\lambda v.v\gets n;\IfO !v \Then x \Else \tr_u (\pred !v))
  \end{mathpar}
\end{definition}
In other words, $\tr_u$ is the function that will return the $i$-th term of the sequence $u$ when given $i$ as an input, or will fail to terminate if passed some index $k\ge|u|$.

\begin{proposition}
  Let $u \in X^*$ and let $n<|u|$.  
  Then it is possible to deduce that
  \[
    \inferrule{\Gamma,s \ts M \converges n,s'}{\Gamma,s \ts \tr_u M \converges u^{(n)},s'}
    \]
  in Idealized Algol.
  \label{PropTr}
\end{proposition}
\begin{proof}
  Induction on $|u|$ and on $n$.
  Since $n<u$, $u$ must be non-empty, of the form $xu'$.

  Suppose $n=0$.  
  Then $u^{(n)}=x$, and we have a derivation of $\Gamma,s\ts \tr_{xu} M\converges x,s'$ from $\Gamma,s\ts M\converges n,s'$ as shown in Figure \ref{FigFirstTermOfSequence}.

  Now suppose that $n=m+1$.  
  Then $(xu)^{(m+1)}=u^{(m)}$.  
  Then we have a derivation of $\Gamma,s\ts \tr_{xu} M \converges u^{(m)},s'$ from $\Gamma,s\ts M\converges n,s'$ in Figure \ref{FigGeneralTermOOfSequence}, using the inductive hypothesis to tell us that we may derive
  \[
    \inferrule*{\Gamma,v,(s'\vert v\mapsto m+1) \ts \pred !v \converges m,(s'\vert v \mapsto m+1)}
    {\Gamma,v,(s'\vert v \mapsto m+1) \ts \tr_u(\pred !v) \mapsto u^{(m)},(s'\vert v\mapsto m+1)}\,.\qedhere
    \]
  \begin{SidewaysFigure}
    \tiny
    \begin{subfigure}{\textheight}
      \centering
      \[
        \inferrule*
        {
          \inferrule*
          {
            \inferrule*
            {
              \inferrule*
              {
                \Gamma,v,(s\vert v \mapsto 0) \ts M \converges 0,(s'\vert v\mapsto 0)
                \\
                \inferrule*
                {
                }
                {
                  \Gamma,v,(s'\vert v \mapsto 0)\ts v \converges v,(s'\vert v\mapsto 0)
                }
              }
              {
                \Gamma,v,(s\vert v\mapsto 0) \ts v \gets M \converges \skipp,(s'\vert v\mapsto 0)
              }
              \\
              \inferrule*
              {
                \inferrule*
                {
                  \inferrule*
                  {
                  }
                  {
                    \Gamma,v,(s'\vert v\mapsto 0) \ts v \converges v,(s'\vert v\mapsto 0)
                  }
                }
                {
                  \Gamma,v,(s'\vert v\mapsto 0) \ts !v \converges 0,(s'\vert v\mapsto 0)
                }
                \\
                \inferrule*
                {
                }
                {
                  \Gamma,v,(s'\vert v\mapsto 0) \ts x \converges x,(s'\vert v\mapsto 0)
                }
              }
              {
                \Gamma,v,(s'\vert v\mapsto 0) \ts \IfO !v \Then x \Else \tr_u(\pred !v) \converges x,(s'\vert v\mapsto 0)
              }
            }
            {
              \Gamma,v,(s\vert v\mapsto 0) \ts v\gets M;\IfO !v \Then x \Else \tr_u(\pred !v) \converges x,(s'\vert v\mapsto 0)
            }
          }
          {
            \Gamma,s\ts \neww (\lambda v.v\gets M;\IfO !v \Then x \Else \tr_u(\pred !v)) \converges x,s'
          }
        }
        {
          \Gamma,s\ts \lambda n.\neww (\lambda v.v\gets n;\IfO !v \Then x \Else \tr_u(\pred !v)) M \converges x,s'
        }
        \]
      \caption{IA derivation that if $M\converges 0$ then $\tr_u M$ converges to the first element of the sequence $u$.}
      \label{FigFirstTermOfSequence}
    \end{subfigure}
    \par\vspace{24pt}
    \begin{subfigure}{\textheight}
    \centering
      \[
        \inferrule*
        {
          \inferrule*
          {
            \inferrule*[sep=5pt]
            {
              \inferrule*
              {
                \inferrule*[rightskip=30pt]
                {
                }
                {
                  \Gamma,v,(s'\vert v\mapsto 0) \ts v \converges v,(s'\vert v\mapsto 0)
                }
                \\\\
                \inferrule*[leftskip=15pt]{}{\Gamma,v,(s\vert v\mapsto 0) \ts M \converges m+1, (s'\vert v\mapsto 0)}
              }
              {
                \Gamma,v,(s\vert v\mapsto 0) \ts v \gets M \converges \skipp,(s'\vert v\mapsto m+1)
              }
              \\
              \inferrule*[sep=5pt]
              {
                \inferrule*
                {
                  \inferrule*
                  {
                  }
                  {
                    \Gamma,v,(s'\vert v\mapsto m+1) \ts v \converges v,(s'\vert v\mapsto m+1)
                  }
                }
                {
                  \Gamma,v,(s'\vert v\mapsto m+1) \ts !v \converges m+1,(s'\vert v\mapsto m+1)
                }
                \\
                \inferrule*
                {
                  \inferrule*
                  {
                    \inferrule*
                    {
                      \inferrule*
                      {
                      }
                      {
                        \Gamma,v,(s'\vert v\mapsto m+1) \ts v \converges v,(s'\vert v\mapsto m+1)
                      }
                    }
                    {
                      \Gamma,v,(s'\vert v\mapsto m+1) \ts !v \converges m+1,(s'\vert v\mapsto m+1)
                    }
                  }
                  {
                    \Gamma,v,(s'\vert v\mapsto m+1) \ts \pred !v \converges m,(s'\vert v\mapsto m+1)
                  }
                }
                {
                  \Gamma,v,(s'\vert v\mapsto m+1) \ts \tr_u(\pred !v) \converges u^{(m)}, (s'\vert v \mapsto m+1)
                }
              }
              {
                \Gamma,v,(s'\vert v\mapsto m+1) \ts \IfO !v \Then x \Else \tr_u(\pred !v) \converges u^{(m)}, (s'\vert v\mapsto m+1)
              }
            }
            {
              \Gamma,v,(s\vert v\mapsto 0) \ts v \gets M;\IfO !v \Then x \Else \tr_u(\pred !v) \converges u^{(m)}, (s'\vert v \mapsto m+1)
            }
          }
          {
            \Gamma,s \ts \neww (\lambda v.v\gets M;\IfO !v \Then x \Else \tr_u(\pred !v)) \converges u^{(m)},s'
          }
        }
        {
          \Gamma,s\ts \lambda n.\new (\lambda v.v\gets n;\IfO !v \Then x \Else \tr_u(\pred !v)) M \converges u^{(m)},s'
        }
        \]
        \caption{IA derivation that if $M\converges m+1$ then $\tr_{u}M$ converges to the $m+1$-th element of the sequence $u$.}
        \label{FigGeneralTermOOfSequence}
    \end{subfigure}
    \par\vspace{24pt}
    \begin{subfigure}{\textheight}
      \[
        \inferrule*
        {
          \inferrule*
          {
            \inferrule*
            {
              \inferrule*
              {
                \inferrule*
                {
                }
                {
                  \Gamma,v,(s\vert v\mapsto k) \ts v \converges v,(s\vert v\mapsto k)
                }
              }
              {
                \Gamma,v,(s\vert v\mapsto k) \ts !v \converges k,(s\vert v\mapsto k)
              }
            }
            {
              \Gamma,v,(s\vert v\mapsto k) \ts \suc !v \converges k+1,(s\vert v\mapsto k)
            }
            \\
            \inferrule*
            {
            }
            {
              \Gamma,v,(s\vert v\mapsto k) \ts v\converges v,(s\vert v\mapsto k)
            }
          }
          {
            \Gamma,v,(s\vert v\mapsto k) \ts v\gets \suc !v \converges \skipp,(s\vert v\mapsto k+1)
          }
          \\
            \inferrule*[right={Prop. \ref{PropTr}}]
            {
              \inferrule*
              {
                \inferrule*
                {
                }
                {
                  \Gamma,v,(s\vert v\mapsto k+1) \ts v\converges v,(s\vert v\mapsto k+1)
                }
              }
              {
                \Gamma,v,(s\vert v\mapsto k+1) \ts !v \converges k+1,(s\vert v\mapsto k+1)
              }
            }
            {
              \Gamma,v,(s\vert v\mapsto k+1) \ts \tr_w !v \converges x,(s\vert v\mapsto k+1)
            }
          }
        {
          \Gamma,v,(s\vert v\mapsto k) \ts v\gets \suc !v;\tr_w !v \converges x,(s\vert v\mapsto k+1)
        }
        \]
      \caption{IA derivation that $(x\mapsto k),v \gets \suc !v;\tr_w !v$ converges to the $k+1$-th term of $w$.}
      \label{FigTheOneForTheLemma}
    \end{subfigure}
    \normalsize
    \caption{Some useful IA derivations}
  \end{SidewaysFigure}
\end{proof}

We start with a small lemma to help us deal with substitution.

@@

>>>>

\begin{example}
  In the category of sets, the \emph{nonempty powerset functor} $\powerset_+$ sends a set $A$ to the set of nonempty subsets of $A$.  
  This functor has the structure of a monad on $\Set$, since we have a natural transformation (union) from $\powerset_+\powerset_+A$ to $\powerset_+A$ and a natural transformation (singleton) from $A$ to$ \powerset_+A$ that obey the diagrams given above.
\end{example}
\begin{example}
  Let $\M$ be a monoidal category and let $x$ be a monoid in $\M$.  
  The \emph{writer monad} $W_x$ on $\M$ is defined by $W_x y = y \tensor x$, with natural transformations
  \begin{mathpar}
    m_y \from y \tensor x \tensor x \to y \tensor x
    \and
    u_y \from y \to y \tensor x
  \end{mathpar}
  given by the monoid structure on $x$.

@@

  Let us look at an example of a monad that arises from an adjunction in this way.  
  Recall that the definition of a monoidal closed category $\M$ is that it admits an adjunction between functors $\blank\tensor w$ and $w\implies\blank$ for any object $w$ of $M$.  
  If $\M$ is a monoidal closed category and $w$ is an object of $\M$, then the functor $S_w$ on $\M$ defined by
  \[
    S_w x = w \implies (x \tensor w)
    \]
  is the composition of these adjoint functors and therefore inherits the structure of a monad, known as the \emph{state monad}.
\end{example}
\begin{example}
  Another example that arises from an adjunction is the \emph{list monad} on $\Set$ that arises from the adjunction between the category $\Set$ of sets and the category of monoids.
  The underlying set of the free monoid on a set $A$ is the set $A^*$ of finite lists of elements of $A$, and the functor $A\mapsto A^*$ inherits a monoid structure where the multiplication $m_A\from (A^*)^* \to A^*$ concatenates a list of lists into a single list and the unit $u_a \from A \to A^*$ forms a list with a single element.  
\end{example}
\begin{example}
  A monad on $\oppcat\C$ is called a \emph{comonad} on $\C$.  
  The carrier of a comonad is still a functor $M\from \C\to\C$, but now the multiplication and unit are natural transformations $M\Rightarrow MM$ and $M\Rightarrow 1$, rather than the other way round.

  An adjunction $\adjunction\C LR\D$ gives rise to a comonad structure on $LR$ in much the same way as it gives rise to a monad structure on $RL$.  
  So, for example, we have the \emph{store comonad} $S_r'$ for any object $r$ of a monoidal closed category $\M$, given by
  \[
    S_r'x = (r \implies x) \tensor r\,.
    \]
\end{example}

@@

\section{Multiplicative natural transformations}

@@

The proof of Proposition \ref{pKleisli} essentially comes down to the following factorization result.  
If $f\from a \to b$ is a morphism in $\Kl_M$, then $f$ may be factorized as
\[
  f = a \xrightarrow{Jf}
  Mb \xrightarrow{\phi_b}
  b\,,
  \]
where we use `$f$' to refer both to the morphism $a\to b$ in $\Kl_M$ and to the underlying morphism $a \to Mb$ in $\C$.
Indeed, if we compute this composite inside $\C$, we get
\[
  a \xrightarrow{f}
  Mb \xrightarrow{u_{Mb}}
  MMb \xrightarrow{M\id}
  MMb \xrightarrow{m_b}
  Mb\,,
  \]
which is equal to $f$ by the coherence conditions on $m$ and $u$.
This means that the Kleisli category may be thought of as being freely generated from the original category $\C$ and an $M$-multiplicative natural transformation $\phi$.

\begin{example}
  The morphisms in the Kleisli category for the nonempty powerset monad $\powerset_+$ on $\Set$ are functions $A \to \powerset_+B$, which can be thought of as nondeterministic functional programs.  
  Given a set $A$, the morphism $\phi_A\from \powerset_+A \to A$ in $\Kl_{\powerset_+}$ can be interpreted as a `nondeterministic choice' function that accepts a nonempty set of elements of $A$ and nondeterministically chooses one of them.
  The factorization then means that the category is freely generated over $\C$ by these nondeterministic choice morphisms.
\end{example}
\begin{example}
  Let $\C$ be a Cartesian closed category and let $z$ be some fixed object of $\C$.  
  Then the Kleisli category for the reader monad $R_z$ on $\C$ is generated over $\C$ by a natural transformation $\phi_y\from (z \to y) \to y$.  
  As a Cartesian closed category, $\C$ may be regarded as being enriched over itself.
  By the enriched Yoneda lemma, then, such a natural transformation is always given by precomposition with some fixed morphism $\ask_z\from 1 \to z$.  
  This means that $\Kl_{R_z}$ is suitable for modelling any situation in which we are generally working in $\C$, but need the ability to request a value of type $z$ (for example, a config file, a piece of user input, a random number or something else that isn't being passed into the function in question).
  \label{ExReaderMonadKleisli}
\end{example}

@@

Having given an overview of the general theory of monads and Kleisli categories, we now examine the relationship between Kleisli categories and Full Abstraction.  From now till the end of the chapter, we fix an base Cartesian closed category $\G$ that admits a denotational semantics of Idealized Algol satisfying Computational Adequacy.  
In addition, we require that $\G$ can be interpreted as being enriched in algebraic partial orders, in such a way that any compact morphism between the denotation of IA types is the denotation of some IA term.
The prototypical example, of course, will be the category of games and visible strategies that we met in Chapter \ref{ChapGames}, but we will not exploit any properties of this model beyond the ones we have already mentioned, mentioning it only in examples where appropriate.

@@

From Proposition \ref{pKleisli}, we know that there is a distinguished natural transformation $\phi_A \from (X \to A) \to A$ in $\G_X$; in particular, we have a morphism
\[
  \phi = \Lambda(\lunit_X;\id_X);\phi_X
  \]
(where $\Lambda(\lunit_X;\id_X)\from 1 \to (X \to X)$ is the interpretation of the $\lambda$-term $\lambda x.x$), which will be the denotation of the term $\ask_X$.
Together with the existing denotational semantics of Idealized Algol within $\G$, this gives us an inductively defined denotational semantics of \IAX within $\G_X$.

@@

  \begin{figure}
    \small
    \begin{mathpar}
      \inferrule*{ }{\Gamma,s\ts c \converges_\epsilon c,s}
      \and
      \inferrule*{\Gamma,s \ts M \converges_u \lambda x.M',s' \\ \Gamma,s' \ts M'[N/x] \converges_v c,s''}{\Gamma,s \ts MN \converges_{u\cat v} c,s''}
      \and
      \inferrule*{\Gamma,s \ts M(\Y M) \converges_u c,s'}{\Gamma,s \ts \Y M \converges_u c,s'}
      \and
      \inferrule*{\Gamma,s\ts M \converges_u n,s'}{\Gamma,s\ts \suc M \converges_u n+1,s'}
      \\\and
      \inferrule*{\Gamma,s\ts M \converges_u n+1,s'}{\Gamma,s\ts \pred M \converges_u n,s'}
      \and
      \inferrule*{\Gamma,s\ts M \converges_u 0,s'}{\Gamma,s\ts \pred M \converges_u 0,s'}
      \and
      \inferrule*{\Gamma,s\ts M \converges_u \skipp,s' \\ \Gamma,s'\ts N \converges_v c,s''}{\Gamma,s \ts M;N \converges_{u\cat v} c,s''}
      \and
      \inferrule*{\Gamma,s\ts M \converges_u \true,s' \\ \Gamma,s' \ts N \converges_v c,s''}{\Gamma,s \ts \If M \Then N \Else P \converges_{u\cat v} c,s''}
      \and
      \inferrule*{\Gamma,s\ts M \converges_u \false,s' \\ \Gamma,s' \ts P \converges_v c,s''}{\Gamma,s \ts \If M \Then N \Else P \converges_{u\cat v} c,s''}
      \and
      \inferrule*{\Gamma,s\ts M \converges_u 0,s' \\ \Gamma,s' \ts N \converges_v c,s''}{\Gamma,s \ts \IfO M \Then N \Else P \converges_{u\cat v} c,s''}
      \and
      \inferrule*{\Gamma,s\ts M \converges_u n+1,s' \\ \Gamma,s' \ts P \converges_v c,s''}{\Gamma,s \ts \IfO M \Then N \Else P \converges_{u\cat v} c,s''}
      \\
      \inferrule{\Gamma,s\ts M\converges c',s' \\ \Gamma,s' \ts N[c'/x]\converges c,s''}{\Gamma,s\ts \lett x=M \inn N \converges c,s''}
      \\
      \inferrule*[right=$x\in\Gamma$]{\Gamma,s\ts E \converges_u n,s' \\ \Gamma,s' \ts V \converges_v x,s''}{\Gamma,s\ts V \gets E \converges_{u\cat v} \skipp,(s''\vert x \mapsto n)}
      \and
      \inferrule*[right={$s'(x)=n$}]{\Gamma,s\ts V \converges_u x,s'}{\Gamma,s\ts !V \converges_u n,s'}
      \and
      \inferrule*{\Gamma,x\from\Var,(s\vert x\mapsto 0)\ts M \converges_u c,(s'\vert x\mapsto n)}{\Gamma,s\ts \neww \lambda x.M \converges_u c,s'}
      \and
      \inferrule*{\Gamma,s\ts E \converges_u n,s' \\ \Gamma,s'\ts V \converges_v \mkvar W R,s'' \\ \Gamma,s'' \ts Wn \converges_w \skipp,s'''}
      {\Gamma,s \ts V\gets E \converges_{u\cat v\cat w} \skipp,s'''}
      \and
      \inferrule*{\Gamma,s\ts V \converges_u \mkvar W R,s' \\ \Gamma,s'\ts R \converges_v n,s''}{\Gamma,s\ts !V \converges_{u\cat v} n,s''}
      \\\and
      \inferrule*[right=$x\in X$]{ }{\Gamma,s\ts \ask_X \converges_x x,s}
    \end{mathpar}
    \caption[Operational semantics for \IAX]{Operational semantics for \IAX.  
    All the rules except the last one are deterministic and may be obtained from the corresponding rules of Idealized Algol by suitably annotating the $\converges$ relation with sequences from $X^*$.}
    \label{FigIaxOpSem}
  \end{figure}

  Closer examination of the rules in Figure \ref{FigIaxOpSem} reveals an alternative, indirect definition of the operational semantics of \IAX.
  Note that each rule from ordinary Idealized Algol takes the form
  \[
    \inferrule{\Gamma,s^{(0)}\ts M_1\converges c_1,s^{(1)} \\ \cdots \\ \Gamma,s^{(n-1)}\ts M_n\converges c_n,s^{(n)}}
    {\Gamma,s^{(0)} \ts M \converges c,s^{(n)}}\,,
    \]
  Here, we have interpreted each IA rule as an infinite scheme of rules ranging over the different terms $M_i,M$ that the rule can apply to.
  We first extend this scheme of rules to a scheme of rules for \IAX, by allowing the $M_i,M$ to range over terms of \IAX.
  We then replace the rule with the new rule
  \[
    \inferrule{\Gamma,s^{(0)}\ts M_1 \converges_{u_1} c_1,s^{(1)} \\ \cdots \\ \Gamma,s^{(n-1)}\ts M_n \converges_{u_n} c_n,s^{(n)}}
    {\Gamma,s^{(0)} \ts M \converges_{u_1 \cat \cdots \cat u_n} c,s^{(n)}}\,,
    \]
  to give us an operational rule for \IAX (if $n=0$, then we treat the empty string $\epsilon$ as the empty concatenation).
  Lastly, we add the rule for the new constant $\ask_X$:
  \[
    \inferrule*[right=$x\in X$]{ }{\Gamma,s \ts \ask_X \converges_x x,s}\,.
    \]
  This rule is the only nondeterministic one in our language, as well as being the only one in which the sequence annotating the $\converges$ symbol at the bottom is not formed from concatenating together the sequences on the top.
\end{definition}

\begin{example}
  If $X=\bC$, then, since $X$ has a single element $a$, a sequence $u=\underbrace{a\cdots a}_n$ of elements of $X$ may be identified with its length $n$.
  In this case, the language \IAX gives us a way to model time complexity, and the term $\ask_X$ may be considered as a subroutine $\sleep\from \com$ whose semantics is to wait for some fixed period of time before continuing.  
  In this case, 
  \[
    \Gamma,s\ts M \converges_n c,s'
    \]
  is interpreted to say that `starting with the store $s$, $M$ converges to $c$ in time $n$, leaving store $s'$'.
\end{example}
\begin{example}
  If $X=\bN$, we can interpret the language \IAX as giving us a way to model user input.  
  When we call $\ask_\bN$, we are asking the user to provide us as a string of text (which we interpret as a binary string and hence as a natural number).  
  Then we interpret the judgement
  \[
    \Gamma,s\ts M \converges_u c,s'
    \]
  as saying that `starting with the store $s$, and the sequence $u$ of pieces of user input, $M$ converges to $c$, leaving store $s'$'.
\end{example}
\begin{example}
  If $X\in\{\bB,\bN\}$, then the language \IAX gives us a way to model nondeterminism, where $\ask_X$ behaves as a \emph{nondeterministic oracle}; i.e., a device that nondeterministically returns an element of $X$.

@@

\begin{definition}
  We inductively define terms in context $\tr_u\from \nat \to X$ of \emph{ordinary deterministic} Idealized Algol for each $u\in X^*$ as follows.
  \begin{mathpar}
    \tr_\epsilon = \lambda n.\Omega
    \and
    \tr_{xu} = \lambda n.\lett y=n \inn \IfO y \Then x \Else \tr_u(\pred y)
  \end{mathpar}
\end{definition}
In other words, $\tr_u$ is the function that will return the $i$-th term of the sequence $u$ when given $i$ as an input, or will fail to terminate if passed some index $k\ge|u|$.

\begin{proposition}
  Let $u \in X^*$ and let $n<|u|$.  
  Then it is possible to deduce that
  \[
    \inferrule{\Gamma,s \ts M \converges n,s'}{\Gamma,s \ts \tr_u M \converges u^{(n)},s'}
    \]
  in Idealized Algol.
  \label{PropTr}
\end{proposition}

We start our soundness proof with a small lemma to help us deal with substitution.

@@

Now, if we have a Kleisli morphism $\sigma\from \oc X \implies A$, then $\sigma\vert_A$ is a nondeterministic strategy for $A$.  
The composition of nondeterministic strategies respects the composition and identity in the Kleisli category; moreover, any two Kleisli morphisms that give rise to the same nondeterministic strategy are intrinsically equivalent.
